---
title: "PML"
output: html_document
date: "2023-07-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Practical Machine Learning Prediction Assignment"

## Introduction
This report discusses the creation of a model to characterize activity based on wearable data using the caret package.

```{r}
library(caret)
```


## Data Acquisition and Preprocessing
Two sets of data are available:  
* training  
* testing  

The data sets are imported as follows:
```{r}
### Import the data

training <- read.csv("pml-training.csv", stringsAsFactors=FALSE, na.strings = c("", "#DIV/0", "NA"))
testing <- read.csv("pml-testing.csv", stringsAsFactors=FALSE, na.strings = c("", "#DIV/0", "NA"))

```

The training data has two issues:  
* NA's throughout  
* The first seven columns are arbitrary  

The data is cleaned by:  
* Removing columns with NA's  
* Removing columns 1 to 7  
* Checking for variables with Near Zero Variance in the remaining data  
    + None of the variables have Near Zero Variance.

```{r}
### Clean the data
training_clean <- training[, colSums(is.na(training)) == 0]
testing_clean <- testing[, colSums(is.na(training)) == 0]

### Remove columns 1-7
training_clean <- training_clean[-c(1:7)]
testing_clean <- testing_clean[-c(1:7)]


### Check for near zero variance
nearZeroVar(training_clean)
```

###  Model Consideration and Testing
Three different models were considered:  
* rpart  
* k nearest neighbors  
* Random Forest

To test the model validity, the training data was partitioned into a train and test set.
For each model, the test set was used to evaluate accuracy.

```{r}
### Partition data

set.seed(3456)
inTrain <- createDataPartition(training_clean$classe, p=0.7, list=FALSE)
training_clean_train <- training_clean[inTrain, ]
training_clean_test <- training_clean[-inTrain, ]
```

For each model, a training model is created using `training_clean_train` data and then tested with the `training_clean_test` data.

The accuracy of the three models differed significantly:

Model Accuracy  
rpart 49.1%  
knn   90.8%  
rf    99.4%  

The RF model is selected for the prediction portion of the assignment.

The final model of the RF model is called to get the OOB estimate for the error rate:
OOB estimate of  error rate: 0.71%

The models were cross-validated via the data partitioning, and also as part of the optimization and internal selection that occurs within the caret package.


Here, the three models are created.
After the RF model is completed, the `finalModel` is called to determine the OOB error rate.

```{r}
### rpart
m_rpart <- train(classe ~ ., data = training_clean_train, method = "rpart")
m_rpart
m_rpart$finalModel
test_output <- predict(m_rpart, training_clean_test[, -53])
mean(training_clean_test[, 53] == test_output)
table(training_clean_test[, 53], test_output)


### knn
m_knn <- train(classe ~ ., data = training_clean_train, method = "knn")
m_knn
m_knn$finalModel
test_output <- predict(m_knn, training_clean_test[, -53])
mean(training_clean_test[, 53] == test_output)
table(training_clean_test[, 53], test_output)

### rf
m_rf <- train(classe ~ ., data = training_clean_train, method = "rf")
m_rf
m_rf$finalModel
test_output <- predict(m_rf, training_clean_test[, -53])
mean(training_clean_test[, 53] == test_output)
table(training_clean_test[, 53], test_output)

```


The final model is called to see the most important variables in the RF model.

```{r}
tmp_2 <- m_rf$finalModel
rfImp <- varImp(m_rf, scale = FALSE)
plot(rfImp, top = 20)

```


A data frame is created to compile the predictions of the three models and compare their output for the final 20-row Test set.

```{r}
N <- 20
compiled_results <- data.frame(
  id = numeric(N),
  rpart = character(N),
  knn = character(N),
  rf = character(N))

compiled_results$id <- 1:20
test_output_final <- predict(m_rpart, testing_clean[, -53])
compiled_results$rpart <- test_output_final
test_output_final <- predict(m_knn, testing_clean[, -53])
compiled_results$knn <- test_output_final
test_output_final <- predict(m_rf, testing_clean[, -53])
compiled_results$rf <- test_output_final

compiled_results
write.table(compiled_results, "compiled_results.csv")

```
